Asymmetric tail-dependence in extreme climatic events affects population-extinction risk

### 2019-08-30 MainText.Rmd
results edited

<!--Q1 : age structured results : linear and nonlinear model with asymmetric copula : Fig. S1-->
<!--Q1 : age structured results : linear and nonlinear model with extreme tail dep. copula : Fig. 1 -->

For single patch linear age-structured models, we found higher extinction risk for moderate left-tail 
dependence (modeled by Clayton copula) in fluctuating vital 
rates compared with the one when vital rates were right tail dependent (modeled by Survival Clayton copula).
It was expected because an extreme "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby can increase the extinction risk of the population. For non-linear age-structured 
models, we got the similar trend for extinction risks with different tail-dependence (left and right).
Non-linear model showed higher extinction risk than its corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{SM-fig_extrisk_agestr_mtd} represents this scenario with two copulas having different 
asymmetric tails but same Kendall's correlation ($\tau = 0.6$). Extinction risks from 
two different tail-dependent copulas were not strinkingly different (e.g. see pair of red and
blue lines in Fig. \ref{SM-fig_extrisk_agestr_mtd}A or B) and we anticipated the difference being partly 
dependent on the degree of tail dependence. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got the similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr_etd}). This result answered (Q1) and showed extinction risk of 
a population could be considered as direct manifestation of tail-dependence property 
between vital rates of different age-class
of that population. <!--discussion later: add Tuljapurkar's approximation-->

<!--Q2 : metapopulation results : density dependence with extreme tail dep. copula-->
<!--For local dispersal : all model summary : Fig. 2, individual model : Fig. S3-S8-->
<!--For global dispersal : all model summary : Fig. S2, individual model : Fig. S9-S14-->

We found density-dependent metapopulation models in stochastic 
environment showed similar patterns for extinction risk as density-independent 
metapopulation model showed in recent study [@Ghosh_copula]. For each density-dependent 
model listed in Table \ref{tab_model_summary}, we observed higher extinction risk for 
the metapopulations when noises across multiple patches were left-tail dependent than the one with right-tail dependence.
But that observation remained valid when the model parameter - here considered as growth rate $(r)$ - of each model was 
sufficiently low so that under-compensatory dynamics (monotonous approach towards
its stable equilibrium) for the deterministic analog of the model was ensured.
We observed right tail dependent noise caused higher extinction 
risk for the metapopulations than left tail dependent noise when the model 
was in over-compensatory regime (oscillatory approach towards stable equilibrium) for higher 
value of $r$. Figs. \ref{fig_extrisk_allmodel_localdisp_etd} and \ref{SM-fig_extrisk_allmodel_localdisp_mtd} show
variation of extinction risk against local dispersal rate ($d$) for six density dependent metapopulation models 
where growth rate $r$ was taken either from under-compensatory or from over-compensatory 
regime. We also got similar results with 
our prediction (see Q2 from \nameref{Introduction}) for global 
dispersal (Figs. \ref{SM-fig_extrisk_allmodel_globaldisp_etd} and \ref{SM-fig_extrisk_allmodel_globaldisp_mtd}). 
Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1_etd} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6_etd}) 
shows how extinction risk changed against local dispersal rate for a whole range
of $r$, whereas Figs. (\ref{SM-fig_ricker_risk_global_disp_scl_1_etd} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6_etd}) 
represents the same but for the case when global dispersal took place between patches. 
Both case showed the signature of 
"rescue-effect" (i.e. by lowering extinction risk) when there was 
a finite dispersal compared to the one with 
$d=0$.

<!--Q3 : metapopulation results : switching near rc for d=0, Fig. 3-->
Following Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1_etd} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6_etd}) we hypothesized that the switching 
between the extinction risk generated by left-tail dependent and right-tail dependent noise 
occurred near the critical parameter $r_c$ (see Q3). For simplicity we only consider 
$d=0$ case, i.e., when there was no dispersal between patches (see \nameref{Methods}). 
Fig. \ref{fig_extrisk_d0} shows higher extinction risk with left-tail dependent noise 
for $r<r_c$ and with right-tail dependent noise for $r>r_c$. $r_c$ was the vertical
line in Fig. \ref{fig_extrisk_d0} indicating the value of $r$ for each model where 
their deterministic analog switches from under- to over-compensatory dynamics. 
Though the switching visually appered far from the theoretical $r_c$ value for Verhulst 
and Austin-Brewer model, they were consistent with our hypothesis as they had comperatively 
smaller range of $r$ than the other models.

# Discussion

@@

 <!--FIG : age structure model results : linear, nonlinear-->
\begin{figure*}[!h]
\textbf{ \hspace{5.4 cm} (A) \hspace{4 cm} (B)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_nonlinear_omg_100.pdf}\\
\caption{Extinction risk from single patch age structured $(A)$ linear ($\Omega = \infty$) and $(B)$ nonlinear ($\Omega = 100$) models against time (shading represents 95\% confidence interval). Fecundity of adults and survival of eggs have either extreme left or extreme right tail dependence, $n_s=10000$. \label{fig_extrisk_agestr_etd}}
\end{center}
\end{figure*}

<!--FIG :  summary : all nonlinear model results for local dispersion : extrisk vs. d-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/allmodel_ext_risk_local_disp_extremenoise.pdf}
\caption{For each of six nonlinear density-dependence models presented in Table \ref{tab_model_summary} with local dispersal
among $n=5$ patches, extinction risk after $n_t = 25$ years for $n_s = 10000$ simulations were calculated.
Extinction risk varies as dispersion-rate ($d$) goes from $0$ to $0.5$. Last two columns represent extinction risk calculated for intrinsic growth rate $(r)$ taken from under-compensatory and over-compensatory regime of deterministic model analog, respectively.\label{fig_extrisk_allmodel_localdisp_etd}}
\end{center}
\end{figure*}

 <!--FIG : all nonlinear model results with d=0, extrisk vs. r-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=10 cm]{./Results/common_legend_extreme_noise.pdf}\\
\includegraphics[width=5 cm]{./Results/ricker_scl_1/ricker_extrisk_d0_extremenoise.pdf}
\includegraphics[width=5 cm]{./Results/hassell_scl_1/hassell_extrisk_d0_extremenoise.pdf}\\
\vspace{-0.6 cm}
\includegraphics[width=5 cm]{./Results/msmith_scl_0.8/msmith_extrisk_d0_extremenoise.pdf}
\includegraphics[width=5 cm]{./Results/pennycuick_scl_1/pennycuick_extrisk_d0_extremenoise.pdf}\\
\vspace{-0.6 cm}
\includegraphics[width=5 cm]{./Results/verhulst_scl_0.6/verhulst_extrisk_d0_extremenoise.pdf}
\includegraphics[width=5 cm]{./Results/austinbrewer_scl_0.6/abrewer_extrisk_d0_extremenoise.pdf}\\
\caption{Transition of higher to lower extinction risk (calculated after $25$ years) for 
the metapopulations affected by extreme left tail dependent noise across
$5$ multiple patches without dispersal ($d=0$), as intrinsic 
growth rate ($r$) goes from under to over-compensatory regime. Opposite trend
was observed with extreme right tail dependent noise.
Error bars represent the standard errors associated with $10000$
simulations for extinction risk computed at each $r$. Dotted line represents the value of $r$
for each model where their deterministic analog switches from under- to over-compensatory dynamics.
\label{fig_extrisk_d0_etd}}
\end{center}
\end{figure*}

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of six density dependence metapopulation models considered for each patch; last three columns indicated equilibrium population value ($P_e^{i}$), critical growth rate ($r_c^{i}$ : transition point from under- to over-compensatory dynamics of deterministic models) and bifurcation parameter ($r_{bf}^{i}$ : transition point from stable to unstable equilibrium of deterministic models), respectively. Superscript $i=1,2,...,n$ indicates states for $i^{th}$ patch.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```

>>>>

<!--Q1 : age structured results : linear and nonlinear model with asymmetric copula-->

For single patch linear age-structured models, we found higher extinction risk for moderate left-tail 
dependence (modeled by Clayton copula) in fluctuating vital 
rates compared with the one when vital rates were right tail dependent (modeled by Survival Clayton copula).
It was expected because an extreme "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby can increase the extinction risk of the population. For non-linear age-structured 
models, we got the similar trend for extinction risks with different tail-dependence (left and right).
Non-linear model showed higher extinction risk than its corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{SM-fig_extrisk_agestr_mtd} represents this scenario with two copulas having different 
asymmetric tails but same Kendall's correlation ($\tau = 0.6$). Extinction risks for
two different tail-dependent vital rates were not strinkingly different (e.g. see pair of red and
blue lines in Fig. \ref{SM-fig_extrisk_agestr_mtd}A or B) and we anticipated the difference being partly 
dependent on the degree of tail dependence. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got the similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr_etd}). This result answered (Q1) and showed extinction risk of 
a population could be considered as direct manifestation of tail-dependence property 
between vital rates of different age-class
of that population. <!--discussion later: add Tuljapurkar's approximation-->

<!--Q2 : metapopulation results : density dependence with extreme tail dep. copula-->

We found density-dependent metapopulation models in stochastic 
environment showed similar patterns for extinction risk as density-independent 
metapopulation model showed in recent study [@Ghosh_copula]. For each density-dependent 
model listed in Table \ref{tab_model_summary}, we observed higher extinction risk for 
the metapopulations when spatially correlated noises were left-tail dependent than the one with right-tail dependence.
But that observation remained valid when the model parameter - here considered as growth rate $(r)$ - was 
sufficiently low so that under-compensatory dynamics (monotonous approach towards
its stable equilibrium) for the deterministic analog of the model was ensured.
We observed right tail dependent noise caused higher extinction 
risk for the metapopulations than left tail dependent noise when the model 
was in over-compensatory regime (oscillatory approach towards stable equilibrium) for higher 
value of $r$. Figs. \ref{fig_extrisk_allmodel_localdisp_etd} and \ref{SM-fig_extrisk_allmodel_localdisp_mtd} show
variation of extinction risk against local dispersal rate ($d$) for six density dependent metapopulation models 
(see Table \ref{tab_model_summary}) with growth rate $r$ considered either from under-compensatory or 
from over-compensatory regime. We also got similar results with our prediction (see Q2 from \nameref{Introduction})
for global dispersal (Figs. \ref{SM-fig_extrisk_allmodel_globaldisp_etd} and \ref{SM-fig_extrisk_allmodel_globaldisp_mtd}). 
Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1_etd} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6_mtd}) 
demonstrate how extinction risk changed against local dispersal rate for a whole range
of $r$, whereas Figs. (\ref{SM-fig_ricker_risk_global_disp_scl_1_etd} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6_mtd}) 
represent the same but for global dispersal among patches. 
Both case showed the signature of "rescue-effect" (i.e. reduced extinction risk for 
finite dispersal compared with $d=0$).

<!--Q3 : metapopulation results : switching near rc for d=0-->
Following Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1_etd} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6_mtd}) we hypothesized that the switching 
between the extinction risk generated by left-tail dependent and right-tail dependent noise 
occurred near the critical parameter $r_c$ (see Q3). For simplicity we only considered 
$d=0$ case, i.e., when there was no dispersal between patches (see \nameref{Methods}). 
Figs. \ref{fig_extrisk_d0_etd} and \ref{SM-fig_extrisk_d0_mtd} demonstrate higher extinction risk 
with left-tail dependent noise for $r<r_c$ and with right-tail dependent noise 
for $r>r_c$. $r_c$ was the vertical line in Figs. \ref{fig_extrisk_d0_etd} and \ref{SM-fig_extrisk_d0_mtd} 
indicating the value of $r$ for each model where their deterministic analog switches 
from under- to over-compensatory dynamics. Though the switching visually appered far away
from the theoretical $r_c$ value for Verhulst and Austin-Brewer model, they were 
consistent with our hypothesis as they had comperatively 
smaller range of $r$ than the other models.

# Discussion\label{Discussion}

@@

 <!--FIG : age structure model results : linear, nonlinear-->
\begin{figure*}[!h]
\textbf{ \hspace{5.4 cm} (A) \hspace{4 cm} (B)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_nonlinear_omg_100.pdf}\\
\caption{Extinction risk against time from single patch age structured $(A)$ linear ($\Omega = \infty$) and $(B)$ nonlinear ($\Omega = 100$) models for $10000$ simulations; shading represents 95\% confidence interval. Fecundity of adults and survival of eggs have either extreme left or extreme right tail dependence. \label{fig_extrisk_agestr_etd}}
\end{center}
\end{figure*}

<!--FIG :  summary : all nonlinear model results for local dispersion : extrisk vs. d-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/allmodel_ext_risk_local_disp_extremenoise.pdf}
\caption{Extinction risks are computed after $25$ years for six nonlinear density-dependence models (Table \ref{tab_model_summary}) with local dispersal among $n=5$ patches and for $10000$ simulations. Extinction risk varies as dispersion-rate ($d$) goes from $0$ to $0.5$. Last two columns represent extinction risk calculated for intrinsic growth rate $(r)$ taken from under-compensatory and over-compensatory regime of deterministic model analog, respectively.\label{fig_extrisk_allmodel_localdisp_etd}}
\end{center}
\end{figure*}

 <!--FIG : all nonlinear model results with extreme noise, d=0, extrisk vs. r-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=10 cm]{./Results/common_legend_extreme_noise.pdf}\\
\includegraphics[width=5 cm]{./Results/ricker_scl_1/ricker_extrisk_d0_extremenoise.pdf}
\includegraphics[width=5 cm]{./Results/hassell_scl_1/hassell_extrisk_d0_extremenoise.pdf}\\
\vspace{-0.6 cm}
\includegraphics[width=5 cm]{./Results/msmith_scl_0.8/msmith_extrisk_d0_extremenoise.pdf}
\includegraphics[width=5 cm]{./Results/pennycuick_scl_1/pennycuick_extrisk_d0_extremenoise.pdf}\\
\vspace{-0.6 cm}
\includegraphics[width=5 cm]{./Results/verhulst_scl_0.6/verhulst_extrisk_d0_extremenoise.pdf}
\includegraphics[width=5 cm]{./Results/austinbrewer_scl_0.6/abrewer_extrisk_d0_extremenoise.pdf}\\
\caption{Transition of higher to lower extinction risk (calculated after $25$ years) for 
the metapopulations affected by extreme left tail dependent noise across
$5$ multiple patches without dispersal ($d=0$), as intrinsic 
growth rate ($r$) goes from under to over-compensatory regime. Extinction risk for 
extreme right tail dependent noise follows the opposite trend.
Error bars represent the standard errors associated with $10000$
simulations for extinction risk computed at each $r$. Dotted line represents the value of $r$
for each model where their deterministic analog switches from under- to over-compensatory dynamics.
\label{fig_extrisk_d0_etd}}
\end{center}
\end{figure*}

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of six density dependence metapopulation models considered for each patch; last three columns indicate equilibrium population value ($P_e^{i}$), critical growth rate ($r_c^{i}$ : transition point from under- to over-compensatory dynamics of deterministic models) and bifurcation parameter ($r_{bf}^{i}$ : transition point from stable to unstable equilibrium of deterministic models), respectively. Superscript $i=1,2,...,n$ indicates index for $i^{th}$ patch.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6),width="1.4 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```
